A Symbolic Decision Procedure for GKAT

### 2024-05-10 notes.tex
FInally, almost, finished normalized bisimulation

% for code
\usepackage{listings}

@@

% operators
\newcommand{\dom}[1]{\mathrm{dom}(#1)}
\newcommand{\cod}[1]{\mathrm{cod}(#1)}
\DeclareMathOperator{\post}{\mathrm{post}}
\newcommand{\reject}{\mathinner{\mathtt{Reject}}}
\newcommand{\accept}{\mathinner{\mathtt{Accept}}}
\DeclareMathOperator*{\bigplus}{\scalerel*{+}{\sum}}

@@

% texts
\newcommand{\tand}{\text{ and }}
\newcommand{\totherwise}{\text{otherwise}}
\newcommand{\tif}{\text{if }}
\newcommand{\tor}{\text{ or }}
\newcommand{\twhere}{\text{ where }}
\newcommand{\tst}{\text{ s.t.\ }}

% typeset
\newcommand{\itemTitle}[1]{\textbf{#1}}
\newcommand{\keyword}[1]{\emph{#1}}

@@

    \affiliation{
    % \position{Position1}
    \department{Department of Computer Science}              %% \department is recommended
    \institution{Boston University}            %% \institution is required
    \city{Boston}
    \state{MA}
    \postcode{02215}
    \country{USA}                    %% \country is recommended
    }
\else
    \author{
        Cheng Zhang \\
        % \and
        % Arthur Azevedo de Amorim \\
        % \and
        % Marco Gaboardi \\
    }
    \date{}
\fi

@@

Given a finite set \(K\), a \emph{symbolic Guarded Kleene Coalgebra with Tests} (sGKCT) \(𝒮 ≜ ⟨S, ϵ, δ⟩\) over a boolean algebra \(ℬ\) consists of a state set \(S\) consists of a accepting boolean \(ϵ\) and a transition function \(δ\):
\begin{mathpar}
    ϵ: S → ℬ, \and
    δ: S → S → K → ℬ,
\end{mathpar}
where for all states \(s ∈ S\), all the booleans are ``disjoint''; namely the conjunction of any two expression from the set \(ϵ(s) + \{δ(s, s', p) ∣ s' ∈ S, p ∈ K\}\) are false.
The ordering on \(𝒮\) is defined pointwise, namely 
\[⟨ϵ₁, δ₁⟩ ≤ ⟨ϵ₂, δ₂⟩ \text{ when } ∀ s, s' ∈ S, ∀ p ∈ K, ϵ₁(s) ≤ ϵ₂(s) \text{ and } δ₁(s, s', p) ≤ δ₂(s, s', p),\]
where the ordering on the right hand side is defined in the boolean algebra.

Intuitively, the elements of boolean is treated as the ``symbolic'' transitions, which can be think of as a set of labels by the classical stone's representation theorem. 
In particular, when the boolean algebra is the free boolean algebra \(\BExp_B\) generated by a finite set \(B\), the labels can be considered as atoms, since \(\BExp_B ≅ 2^{\At_B}\). 
Then we can explain the intuition for the definition of GKCT
\begin{itemize}
    \item \(ϵ(s)\) can be thought of as a set of labels that is accepted by the state \(s\).
    \item \(δ(s, s', p)\) denotes all the labels that can transition from \(s\) to \(s'\) while executing \(p\).
\end{itemize}
Then the labels that is not in either operations are implicitly rejecting.

Given a sGKCT \(𝒮 ≜ ⟨S, ϵ, δ⟩\) and a state \(s ∈ S\), we use the notation \(s ⇒ b\) to denote that \(ϵ(s) ≥ b\), and we use the notation \(s \transvia{b ∣ p} s'\) to denote that \(δ(s)₃(s', p) ≥ b\),
where the ordering is the typical ordering in boolean algebra.

The symbolic derivatives for GKAT forms a GKCT where the states are represented by GKAT expression \(\GKAT_{K, B}\), and the boolean algebra is the free boolean algebra \(\BExp_B\) over the test alphabet \(B\):
\begin{mathpar}
    ϵ: \GKAT_{K, B} → \BExp_B; \and  
    δ: \GKAT_{K, B} → \GKAT_{K, B} → K → \BExp_B.
\end{mathpar} 
the accepting map \(ϵ\) and transition map \(δ\) can be defined by the smallest maps that satisfy the following rules:
\begin{mathpar}
    \inferrule{\\}{b ⇒ b} \and 
    \inferrule{\\}{p \transvia{1 ∣ p} 1} \and 
    \inferrule{e ⇒ a}{e +_b f ⇒ b a} \and 
    \inferrule{f ⇒ a}{e +_b f ⇒ b̄ a} \and 
    \inferrule{e \transvia{a ∣ p} e'}{e +_b f \transvia{b ∧ a ∣ p} e'} \and 
    \inferrule{f \transvia{a ∣ p} f'}{e +_b f \transvia{b̄ ∧ a ∣ p} f'} \and 
    \inferrule{e ⇒ a \\ f ⇒ b}{e ⋅ f ⇒ a ∧ b} \and 
    \inferrule{e ⇒ a \\ f \transvia{b ∣ p} f'}{e ⋅ f \transvia{a ∧ b ∣ p} f'} \and 
    \inferrule{e \transvia{b ∣ p} e'}{e ⋅ f \transvia{b ∣ p} e' ⋅ f} \and 
    \inferrule{\\}{e^{(b)} ⇒ b̄} \and 
    \inferrule{e \transvia{a ∣ p} e'}{e^{(b)} \transvia{b ∧ a ∣ p} e' ⋅ {e}^{(b)}}
\end{mathpar}

@@

\section{On the Fly Normalization}

Normalization of a GKAT automaton will remove all the dead state, which are states that cannot lead to an accepting transitions. 
However, this process is hard to do on-the-fly, mainly because of the difficulty of handling infinite loops.
\begin{example}
    We can consider a simple extension to a BFS algorithm, where we will mark a state as either live, dead, or searching. 
    A state is marked as ``searching'' when it has been explored but not yet known to be dead or alive yet,
    and when a state loops back to a searching state, then we can declare the detection of a infinite loop, hence marking the transition dead.
    We consider the following coalgebra over a two element set \(\{1, 2\}\):
    \[
    \begin{tikzcd}[row sep=15px]
        1 \ar[bend right, swap]{r}{b ∣ p} \ar[Rightarrow, swap]{d}{b̄}& 2 \ar[bend right, swap]{l}{b ∣ p}\\  
        \ &  
    \end{tikzcd}
    \]
    It is clear that both states \(1\) and \(2\) are live,
    since \(1\) is accepting and \(2\) have a transition to \(1\).
    However, if we run our search algorithm: to determine whether \(1\) is live, we will need to compute the backtrack of both of its transitions. 
    We will compute the backtrack from \(b\) first, which computes the liveness of \(2\). Thus we will mark \(1\) as searching. Yet state \(2\) only have one transition to \(1\), because \(1\) is marked as searching, the algorithm will think a infinite loop is discovered and mark \(2\) as dead, which doesn't align with the reality.
\end{example}

Thus, it is crucial to resolve these loops when we a designing a on-the-fly algorithm.
In fact, we can utilize a common graph theory technique called \emph{graph condensation} to remove the loops, where it treat all thee strong connected component as a vertex as a new graph, and the edges between the components are preserved. 
The resulting graph is a directed acyclic graph (DAG), then we can utilize a simple search to find the live state.

\begin{definition}[Graph For sGKCT]
    Given a sGKCT \(𝒮 ≜ ⟨S, δ, ϵ⟩\), we can construct a (potentially infinite) graph, where the vertices are \(S\) and the edge \((s, s')\) is in the graph when  connected there exists a \(p ∈ K\), s.t. \(δ(s, s', p) ≠ 0\)
\end{definition}

We will also recall the definition of strongly connected components 
\begin{definition}[Walk and strongly connected components]
    A walk from vertex \(s₁\) to vertex \(s₂\) in a graph is a sequence of vertices \([s₁, s₂, …, sₙ]\), s.t. for all \(i\), \((sᵢ, s_{i + 1})\) are edges in the graph.
    A strongly connected component is a equivalence class on the vertices, where two vertices \(s₁\) and \(s₂\) are equivalent when there is a walk from \(s₁\) to \(s₂\) and there is a walk from \(s₂\) to \(s₁\).
\end{definition}

And in fact strongly connected components has strong connections with liveness of states.

\begin{lemma}
    By induction on the length of the walk, if there is a walk from state \(s₁\) and \(s₂\) then if \(s₂\) is live and \(s₁\) is live. 
    Thus by definition of strongly connected component, a strongly connected component is live if and only if any of the state in the component is live. 
    And consequently, all the states in a strongly connected component are either all dead or all live.
\end{lemma}

To identify the strongly connected components, we can run the Tarjan's algorithm along side the normal Hopcroft and Karp's bisimulation algorithm.
The Tarjan's algorithm is a one pass DFS based algorithm where it assigns each state a \emph{low-link value}, at the end of the algorithm, all the algorithm with the same low-link value will be in the same strongly connected components. 
Thus instead of marking each state with dead or live, we will mark each state with a low-link value, and each low-link value is marked with either dead or alive.
This kind of mappings are enabled by the fact that all states in a strongly connected components are either all dead or all live.

\section{The Algorithm}

FIXME: proof a theorem that if the transformation shares boolean with the original program, then the decision procedure is nearly linear time.

>>>>

% for code
\usepackage{listings}
\lstset{basicstyle=\footnotesize\ttfamily,breaklines=true}
\lstset{framextopmargin=50pt,frame=bottomline}
\lstset{language=caml, escapeinside={[*}{*]}}

@@

% operators
\newcommand{\dom}[1]{\mathrm{dom}(#1)}
\newcommand{\cod}[1]{\mathrm{cod}(#1)}
\DeclareMathOperator{\post}{\mathrm{post}}
\DeclareMathOperator{\Img}{\mathrm{\mathbf{Img}}}
\newcommand{\reject}{\mathinner{\mathtt{Reject}}}
\newcommand{\accept}{\mathinner{\mathtt{Accept}}}
\DeclareMathOperator*{\bigplus}{\scalerel*{+}{\sum}}
\newcommand{\closSim}{\mathrel{\overline{\sim}}}
\DeclareMathOperator{\norm}{\mathrm{norm}}

@@

@@

    \affiliation{
    % \position{Position1}
    \department{Department of Computer Science}              %% \department is recommended
    \institution{Boston University}            %% \institution is required
    \city{Boston}
    \state{MA}
    \postcode{02215}
    \country{USA}                    %% \country is recommended
    }
\else
    \author{
    }
    \date{}
\fi

@@

Given a finite set \(K\), a \emph{symbolic Deterministic Kleene Coalgebra with Tests} (sDKCT) \(𝒮 ≜ ⟨S, ϵ, δ⟩\) over a boolean algebra \(ℬ\) consists of a state set \(S\) consists of a accepting boolean \(ϵ\) and a transition function \(δ\):
\begin{mathpar}
    ϵ: S → ℬ, \and
    δ: S → S → K → ℬ,
\end{mathpar}
where for all states \(s ∈ S\), all the booleans are ``disjoint''; namely the conjunction of any two expression from the set \(ϵ(s) + \{δ(s, s', p) ∣ s' ∈ S, p ∈ K\}\) are false.
The ordering on \(𝒮\) is defined pointwise, namely 
\[⟨ϵ₁, δ₁⟩ ≤ ⟨ϵ₂, δ₂⟩ \text{ when } ∀ s, s' ∈ S, ∀ p ∈ K, ϵ₁(s) ≤ ϵ₂(s) \text{ and } δ₁(s, s', p) ≤ δ₂(s, s', p),\]
where the ordering on the right hand side is defined in the boolean algebra.

Intuitively, the elements of boolean is treated as the ``symbolic'' transitions, which can be think of as a set of labels by the classical stone's representation theorem. 
In particular, when the boolean algebra is the free boolean algebra \(\BExp_B\) generated by a finite set \(B\), the labels can be considered as atoms, since \(\BExp_B ≅ 2^{\At_B}\). 
Then we can explain the intuition for the definition of DKCT
\begin{itemize}
    \item \(ϵ(s)\) can be thought of as a set of labels that is accepted by the state \(s\).
    \item \(δ(s, s', p)\) denotes all the labels that can transition from \(s\) to \(s'\) while executing \(p\).
\end{itemize}
Then the labels that is not in either operations are implicitly rejecting.

Given a sDKCT \(𝒮 ≜ ⟨S, ϵ, δ⟩\) and a state \(s ∈ S\), we use the notation \(s ⇒ b\) to denote that \(ϵ(s) ≥ b\), and we use the notation \(s \transvia{b ∣ p} s'\) to denote that \(δ(s)₃(s', p) ≥ b\), where the ordering is the typical ordering in boolean algebra.

The symbolic derivatives for GKAT forms a DKCT where the states are represented by GKAT expression \(\GKAT_{K, B}\), and the boolean algebra is the free boolean algebra \(\BExp_B\) over the test alphabet \(B\):
\begin{mathpar}
    ϵ: \GKAT_{K, B} → \BExp_B; \and  
    δ: \GKAT_{K, B} → \GKAT_{K, B} → K → \BExp_B.
\end{mathpar} 
the accepting map \(ϵ\) and transition map \(δ\) can be defined by the smallest maps that satisfy the following rules:
\begin{mathpar}
    \inferrule{\\}{b ⇒ b} \and 
    \inferrule{\\}{p \transvia{1 ∣ p} 1} \and 
    \inferrule{e ⇒ a}{e +_b f ⇒ b a} \and 
    \inferrule{f ⇒ a}{e +_b f ⇒ b̄ a} \and 
    \inferrule{e \transvia{a ∣ p} e'}{e +_b f \transvia{b ∧ a ∣ p} e'} \and 
    \inferrule{f \transvia{a ∣ p} f'}{e +_b f \transvia{b̄ ∧ a ∣ p} f'} \and 
    \inferrule{e ⇒ a \\ f ⇒ b}{e ⋅ f ⇒ a ∧ b} \and 
    \inferrule{e ⇒ a \\ f \transvia{b ∣ p} f'}{e ⋅ f \transvia{a ∧ b ∣ p} f'} \and 
    \inferrule{e \transvia{b ∣ p} e'}{e ⋅ f \transvia{b ∣ p} e' ⋅ f} \and 
    \inferrule{\\}{e^{(b)} ⇒ b̄} \and 
    \inferrule{e \transvia{a ∣ p} e'}{e^{(b)} \transvia{b ∧ a ∣ p} e' ⋅ {e}^{(b)}}
\end{mathpar}

@@

% \section{On the Fly Normalization}

% Normalization of a GKAT automaton will remove all the dead state, which are states that cannot lead to an accepting transitions. 
% However, this process is hard to do on-the-fly, mainly because of the difficulty of handling infinite loops.
% \begin{example}
%     We can consider a simple extension to a BFS algorithm, where we will mark a state as either live, dead, or searching. 
%     A state is marked as ``searching'' when it has been explored but not yet known to be dead or alive yet,
%     and when a state loops back to a searching state, then we can declare the detection of a infinite loop, hence marking the transition dead.
%     We consider the following coalgebra over a two element set \(\{1, 2\}\):
%     \[
%     \begin{tikzcd}[row sep=15px]
%         1 \ar[bend right, swap]{r}{b ∣ p} \ar[Rightarrow, swap]{d}{b̄}& 2 \ar[bend right, swap]{l}{b ∣ p}\\  
%         \ &  
%     \end{tikzcd}
%     \]
%     It is clear that both states \(1\) and \(2\) are live,
%     since \(1\) is accepting and \(2\) have a transition to \(1\).
%     However, if we run our search algorithm: to determine whether \(1\) is live, we will need to compute the backtrack of both of its transitions. 
%     We will compute the backtrack from \(b\) first, which computes the liveness of \(2\). Thus we will mark \(1\) as searching. Yet state \(2\) only have one transition to \(1\), because \(1\) is marked as searching, the algorithm will think a infinite loop is discovered and mark \(2\) as dead, which doesn't align with the reality.
% \end{example}

% Thus, it is crucial to resolve these loops when we a designing a on-the-fly algorithm.
% In fact, we can utilize a common graph theory technique called \emph{graph condensation} to remove the loops, where it treat all thee strong connected component as a vertex as a new graph, and the edges between the components are preserved. 
% The resulting graph is a directed acyclic graph (DAG), then we can utilize a simple search to find the live state.

% \begin{definition}[Graph For sDKCT]
%     Given a sDKCT \(𝒮 ≜ ⟨S, δ, ϵ⟩\), we can construct a (potentially infinite) graph, where the vertices are \(S\) and the edge \((s, s')\) is in the graph when  connected there exists a \(p ∈ K\), s.t. \(δ(s, s', p) ≠ 0\)
% \end{definition}

% We will also recall the definition of strongly connected components 
% \begin{definition}[Walk and strongly connected components]
%     A walk from vertex \(s₁\) to vertex \(s₂\) in a graph is a sequence of vertices \([s₁, s₂, …, sₙ]\), s.t. for all \(i\), \((sᵢ, s_{i + 1})\) are edges in the graph.
%     A strongly connected component is a equivalence class on the vertices, where two vertices \(s₁\) and \(s₂\) are equivalent when there is a walk from \(s₁\) to \(s₂\) and there is a walk from \(s₂\) to \(s₁\).
% \end{definition}

% And in fact strongly connected components has strong connections with liveness of states.

% \begin{lemma}
%     By induction on the length of the walk, if there is a walk from state \(s₁\) and \(s₂\) then if \(s₂\) is live and \(s₁\) is live. 
%     Thus by definition of strongly connected component, a strongly connected component is live if and only if any of the state in the component is live. 
%     And consequently, all the states in a strongly connected component are either all dead or all live.
% \end{lemma}

% To identify the strongly connected components, we can run the Tarjan's algorithm along side the normal Hopcroft and Karp's bisimulation algorithm.
% The Tarjan's algorithm is a one pass DFS based algorithm where it assigns each state a \emph{low-link value}, at the end of the algorithm, all the algorithm with the same low-link value will be in the same strongly connected components. 
% Thus instead of marking each state with dead or live, we will mark each state with a low-link value, and each low-link value is marked with either dead or alive.
% This kind of mappings are enabled by the fact that all states in a strongly connected components are either all dead or all live.

\section{Normalized Bisimulation}

Bisimilarity implies language sematnical equivalence \[s ∼ t ⟹ ⟦s⟧ = ⟦t⟧\], however the converse doesn't hold unless we are working in a \emph{normal} GKAT coalgebra.
It is well known that two states in a normal GKAT coalgebra if and only if they have the same language semantics \[s ∼ t ⟺ ⟦s⟧ = ⟦t⟧.\]
However, instead of normalizing automatons, we can in fact, check for normality on the fly, this is important to design a derivatives based decision algorithm.

We will use a notion of normalized bisimulation, where we will enlarge the bisimulation by dead states, since all the dead states are trace equivalent.
Operationally, when we find a discrepancy during the bisimulation algorithm we will check whether the states causing the discrepancy are dead.
We can prove the correctness of this algorithm by proving that they are equivalent to normalize and then computing bisimulation.
Let \(G\) be the functor defining GKAT coalgebra: \[G ≜ (2 + (-) × K)^{\At}.\]
Given two DKCT \(𝒮, 𝒯\), and a relation on these two expression \({∼} ⊆ S × T\), we can define \({\closSim} ⊆ (2 + S × K) × (2 + T × K)\) as the smallest relation generated by the following rules: for all \(p, q ∈ K\) and states \(s' ∈ S\), \(t' ∈ T\):
\begin{mathpar}
    \inferrule[Bisim-0]{\\}{0 \closSim 0}
    \and 
    \inferrule[Bisim-1]{\\}{1 \closSim 1}
    \and 
    \inferrule[Bisim-Trans]{s' ∼ t'}{(s', p) \closSim (t', p)}
    \and
    \inferrule[Norm-Left]{s' \text{ is dead}}{(s', p) \closSim 0}
    \and 
    \inferrule[Norm-Right]{t' \text{ is dead}}{0 \closSim (t', p)}
    \and
    \inferrule[Norm-Both]{s', t' \text{ are both dead}}{(s', p) \closSim (t', q)}
\end{mathpar}
Intuitively, \({\closSim}\) enlarges the relation \(∼\) for \emph{results} of a transition. for example, it allows transition to dead states to equivalent to rejection, and so on.
This extension enables ``on-the-fly normalization'', where to determine whether \(s \closSim t\), we can simply first check for bisimulation using the bisimulation rules; and if all failed, we can check the normalization rules.

We call the bisimulation algorithm with on-the fly normalization ``\emph{normalized bisimulation}''.
This concept can be summed up nicely using diagrams.
For a relation \(∼\), then there is a straight forward embedding \((2 + {∼} × K) ↪ {\closSim}\):
This embedding generates all the elements that is formed by the bisimulation rules of \(\closSim\),
also note that \(∼\) is not restricted to live states, but dead states as well; this means that if two dead states follows the bisimulation rules, then we don't need to fall back to the normalization rules.
Since detecting whether a state is dead requires iterating through all its predecessors, we can use this property to forgo dead state detection until all the bisimulation rules failed.

The aforementioned embedding \((2 + {∼} × K) ↪ {\closSim}\) can be point-wise lifted to \(G(∼) ↪ (\At → {\closSim})\), thus our normalized algorithm can be seen as finding a normalized bisimulation \(∼\) that satisfy the following diagram:
\[
    \begin{tikzcd}[column sep=1.5cm, row sep=1cm]
        S \ar[swap]{d}{δ_𝒮}
            & ∼  \ar[dashed, swap]{d}{∃} \ar[swap]{l}{π₁} \ar{r}{π₂}
            & T \ar{d}{δ_𝒯} \\  
        (2 + S × K)^\At 
            & ({\closSim}^\At) 
                \ar[swap]{l}{{π₁}^\At} \ar{r}{{π₂}^\At}
            & (2 + T × K)^\At
    \end{tikzcd}
\]
Notice, unlike \(G(∼)\), the map \({∼} ↦ (\closSim)^\At\) is not a functor,
since it is only defined on relations.

\begin{example}
    Let's consider a example with a single primitive test \(B ≜ \{b\}\),
    thus atoms are defined as \(\At_B = \{\{b\}, ∅\}\).
    We focus on two states in the bisimulation \(s\) and \(t\).
    with their respective transition defined as follows 
    \begin{align*}
            δ_𝒮(s, \{b\}) & ≜ 0   
                & δ_𝒮(t, \{b\}) & ≜ (t', p) \\  
            δ_𝒮(s, ∅) & ≜ (s', p) 
                & δ_𝒮(t, ∅) & ≜ (t', p) 
    \end{align*}
    It is quite clear \(s\) and \(t\) will not be bisimular, however, there can be a normalized bisimulation relation between them, provided that \(s' ∼ t'\) and \(t'\) is dead: 
    \[
    \begin{tikzcd}[column sep=1.5cm, row sep=1cm]
        s \ar[swap, mapsto]{d}{δ_𝒮}
            & (s, t)  \ar[dashed, swap, mapsto]{d}{∃} \ar[swap, mapsto]{l}{π₁} \ar[mapsto]{r}{π₂}
            & t \ar[mapsto]{d}{δ_𝒯} \\  
        \begin{gathered}
            \{b\}↦ 0 \\  
            ∅ ↦ (s', p)
        \end{gathered} 
            & \begin{gathered}
                \{b\}↦ 0, (t', p) \\  
                ∅ ↦ (s', p), (t', p)
            \end{gathered} 
                \ar[swap, mapsto]{l}{{π₁}^\At} \ar[mapsto]{r}{{π₂}^\At}
            & \begin{gathered}
                \{b\} ↦ (t', p) \\  
                ∅ ↦(t', p)
            \end{gathered}
    \end{tikzcd}
    \]
    Notice that \(0 \closSim (t', p)\) because \(t'\) is dead, 
    and \((s', p) \closSim (t', p)\) because we assume that \(s' ∼ t'\). 
    Furthermore, if both \(s'\) and \(t'\) are dead, we can establish \(s' \closSim t'\) using either the \textnormal{\textsc{Bisim-Trans}} rule or \textnormal{\textsc{Norm-Both}} rule, and we are free to pick the more efficient check in our implementation.
\end{example}

\begin{theorem}[Universiality]
    Given a normalized bisimulation relation \({∼} ⊆ S × T\)
    \[
        \begin{tikzcd}[column sep=1.5cm, row sep=1cm]
            S \ar[swap]{d}{δ_𝒮}
                & ∼  \ar[dashed, swap]{d}{h} \ar[swap]{l}{π₁} \ar{r}{π₂}
                & T \ar{d}{δ_𝒯} \\  
            (2 + S × K)^\At 
                & (\closSim)^\At 
                    \ar[swap]{l}{{π₁}^\At} \ar{r}{{π₂}^\At}
                & (2 + T × K)^\At
        \end{tikzcd}
    \] 
    Then \(h\) is unique, and defined as follows: 
    \begin{align*}
        h & : {∼} → {\closSim}^\At \\
        h & (s, t, α) ≜ (δ_𝒮(s, α), δ_𝒯(s, α))
    \end{align*}
\end{theorem}

\begin{proof}
    The commutativity can be verified by computation. We only need to show uniqueness.

    The above commutative diagram implies the commutativity of the following diagram:
    \[
        \begin{tikzcd}
            & {∼} \ar{d}{h} \ar[swap]{ddl}{δ_𝒮 ∘ π₁} \ar{ddr}{δ_𝒯 ∘ π₂}& \\
            & (\closSim)^\At \ar[hook]{d}{i} & \\  
            (2 + S × K)^\At & 
                (2 + S × K)^\At × (2 + T × K)^\At
                    \ar{l}{π₁^\At} \ar[swap]{r}{π₂^\At}
                & (2 + T × K)^\At
        \end{tikzcd}
    \] 
    Since \((-)^{\At}\) is a right adjoint and right adjoints preserve limits, \((2 + S × K)^\At × (2 + T × K)^\At\) is the product of \((2 + T × K)^\At\) and \((2 + T × K)^\At\). 
    Hence, \(h ∘ i\) is the unique function to make the above diagram commute. 
    Assume we can replace \(h\) with \(g\) while preserving the commutativity of the diagram: 
    \begin{align*}
        & i ∘ h  = i ∘ g & \text{\((2 + S × K)^\At × (2 + T × K)^\At\) is a product.} \\  
        ⟹ {}& h = g & \text{\(i\) is a embedding, hence left cancelable.}
    \end{align*}
\end{proof}

\begin{definition}
    Given a DKCT \(𝒮 = ⟨S, δ⟩\), a \emph{sub-DKCT} \(𝒮' = ⟨S', δ|_{S'}⟩\) is a coagebra induced by \(S' ⊆ S\), a sub-DKCT can be characterized as a pullback square:
    \[
        \begin{tikzcd}
            S' \ar[hook]{r}{i} \ar{d}{δ|_{S'}} & S \ar{d}{δ} \\  
            G(S') \ar[hook]{r}{i} & G(S)
        \end{tikzcd}
    \]
    By the commutativity of this diagram \(i\) is also a DKCT homomorphism.
    A \emph{principle sub-DKCT} of state \(s ∈ S\), denoted \(⟨s⟩\) is the smallest sub-DKCT that contains \(s\). 
    We can show that the principle DKCT of \(s\) exactly characterize the DKCT induced by the reachable state of \(s\).
\end{definition}

%%%% NOTE: The universiality is not true, since 𝒯 will need to have the same transition structure as ⟨s⟩. 
%%%%
% \begin{definition}
%     a \emph{principle sub-DKCT} of \(𝒮\) is a ``free'' sub-DKCT generated by one element \(s\), namely, given any sub-DKCT \(𝒯\) and a element \(t ∈ 𝒯\), there is a unique DKCT homomorphism \(h: ⟨s⟩ → 𝒯\), s.t. \(h(s) = t\).

%     We denote the principle sub-DKCT of \(⟨s⟩_{𝒮}\), and we will omit the subscript \(𝒮\) if it is clear from context.
% \end{definition}

% \begin{lemma}
%     The principle sub-DKCT \(⟨s⟩_𝒮\) always exists, and it is the smallest sub-DKCT that contain \(s\).
% \end{lemma}

% \begin{proof}
%     \(⟨s⟩_𝒮\) can be constructed as the set of all the reachable state of \(s\) in \(𝒮\), specifically it is the smallest set that is generated by the following rules:
%     \begin{mathpar}
%         \inferrule{\\}{s ∈ ⟨s⟩_𝒮} \and
%         \inferrule{s₁ ∈ ⟨s⟩_𝒮 \\ ∃ α, δ(s₁, α) = (s₂, p)}{s₂ ∈ ⟨s⟩_𝒮}
%     \end{mathpar}
%     then the homomorphism \(h: ⟨s⟩ → 𝒯\) can be inductively defined as follows:
%     \begin{align*}
%         h(s) & ≜ t \\ 
%         h(s₂) & ≜ h(δ(s₁)) 
%     \end{align*}
% \end{proof}

\begin{lemma}
    A state \(s\) is live if and only if there exists a accepting state in \(⟨s⟩\);  
    and a state \(s\) is dead if and only if there is no accepting state in \(⟨s⟩\).
\end{lemma}

\begin{corollary}
    \(s\) is dead if and only if \(⟨s⟩\) is all dead.
\end{corollary}

\begin{proof}
    Take any state \(s' ∈ ⟨s⟩\), we can construct the DKCT \(⟨s'⟩\).
    Since \(⟨s'⟩\) is the smallest sub-DKCT that contains \(s'\), and \(s' ∈ ⟨s⟩\), thus \(⟨s'⟩ ⊆ ⟨s⟩\).
    Finally, because there is no accepting state in \(⟨s⟩\), there cannot be any accepting state in \(⟨s'⟩\) thus \(s'\) is also dead.
\end{proof}

In fact the normalization operation commutes with principle sub-DKCT. This result is intuitive as taking all the reachable states and then normalize is the same as normalize and then take all the reachable state.

\begin{lemma}
    sub-DKCT preserves principle sub-DKCT, that is, given \(𝒮' ⊑ 𝒮\) and \(s ∈ 𝒮'\), then \[⟨s⟩_{𝒮'} = ⟨s⟩_{𝒮}.\]
\end{lemma}

\begin{proof}
    First by definition, \(s ∈ ⟨s⟩_{𝒮}\),
    then we show that \(⟨s⟩_{𝒮}\) is the smallest sub-DKCT of \(𝒮'\), that contains \(s\) in \(𝒮\):

    Because \(s ∈ ⟨s⟩_{𝒮'}\) and \(⟨s⟩_{𝒮'} ⊑ 𝒮' ⊑ 𝒮\), 
    by minimality of \(⟨s⟩_{𝒮}\): \[⟨s⟩_{𝒮} ⊑ ⟨s⟩_{𝒮'} ⊑ 𝒮',\] that is, \(⟨s⟩_{𝒮}\) is a sub-DKCT of \(𝒮'\).

    Given any \(𝒯  ⊑ 𝒮'\) and \(s ∈ 𝒯\), we have \(𝒯 ⊑ 𝒮' ⊑ 𝒮\),
    thus by minimality of \(⟨s⟩_{𝒮}\): \(⟨s⟩_{𝒮} ⊑ 𝒯\).
    In other word, \(⟨s⟩_{𝒮}\) is the the minimal sub-algebra of \(𝒮'\) that contains \(s\), reaching desired conclusion.
\end{proof}

\begin{lemma}
    sub-DKCT preserves liveness, that is given \(𝒮' ⊑ 𝒮\) and \(s ∈ 𝒮'\), 
    \[s \text{ is live in \(𝒮'\)} ⟺ s \text{ is live in \(𝒮\)}\]
\end{lemma}
\begin{proof}
    recall that sub-DKCT preserves principle sub-DKCT, namely \(⟨s⟩_{𝒮'} = ⟨s⟩_{𝒮}\), thus 
    \begin{align*}
        s \text{ is live in \(𝒮'\)} 
        & ⟺ ⟨s⟩_{𝒮'} \text{ contains an accepting state} \\
        & ⟺ ⟨s⟩_{𝒮} \text{ contains an accepting state} \\
        & ⟺ s \text{ is live in \(𝒮\)} 
    \end{align*}
\end{proof}

\begin{lemma}
    \(\norm\) is a functor, and it preserves inclusion, that is 
    \[𝒮 ⊑ 𝒯 ⟹ \norm(𝒮) ⊑ \norm(𝒯).\]
\end{lemma}

\begin{lemma}
    Given a DKCT \(𝒮\), and two sub-DKCT \(𝒮₁ ⊑ 𝒮\) and \(𝒮₂ ⊑ 𝒮\), \(𝒮₁ = 𝒮₂\) if and only if their carrier set is equal.
\end{lemma}

\begin{lemma}
    Given a DKCT \(𝒮\), its dead states induces a sub-DKCT.
\end{lemma}

\begin{theorem}
    Given a DKCT \(𝒮\) and a live state \(s ∈ 𝒮\):
    \[⟨s⟩_{\norm(𝒮)} = \norm(⟨s⟩_{𝒮}).\]
\end{theorem}

\begin{proof}
    We need to show \(\norm(⟨s⟩_{𝒮})\) is the smallest sub-DKCT in \(\norm(𝒮)\) that contains \(s\).

    First, \(\norm(⟨s⟩_{𝒮})\) contains \(s\) because \(s\) is live. 
    Second, take any \(𝒯 ⊑ \norm(𝒮)\), we will use \(\overline{𝒯}\) to denote the smallest sub-DKCT of \(𝒮\) that contains all the states in \(𝒯\). 
    Since \(𝒯\) is a sub-DKCT of \(\norm(𝒮)\), \(\overline{𝒯} ∖ 𝒯\) can only contain dead states, thus \(\norm(\overline{𝒯}) = 𝒯\).
    \begin{align*}
        & ⟨s⟩_{𝒮} ⊑ \overline{𝒯} & \text{\(\overline{𝒯}\) contains \(s\)} \\  
        ⟹ {}& \norm(⟨s⟩_{𝒮}) ⊑ \norm(\overline{𝒯}) & \text{monotonicity of \(\norm\)} \\  
        ⟹ {}& \norm(⟨s⟩_{𝒮}) ⊑ 𝒯 & \norm(\overline{𝒯}) = 𝒯
    \end{align*}
\end{proof}

We will show this by establishing the equivalence between ``normalized bisimulation'' and ``bisimulation in normalized coalgebra''.
Given a GKAT coalgebra \(𝒮\), we can normalize it via removing the dead states and reroute all the transitions to dead state as direct rejections.
Similarly, any relation \({∼} ⊆ 𝒮 × 𝒯\) can be normalized by filtering out all the states that are live:
\[\norm(∼) ≜ \{(s, t) ∣ \text{both \(s\) and \(t\) are live states}\}.\]
Thus there is a natural embedding \(\norm(∼) ↪ {∼}\) by inclusion. 
Similarly, there is a embedding \(G(\norm(∼)) → (\closSim)^\At\) by lifting the following embedding point-wise via \((-)^\At\):
\begin{align*}
    2 + {\norm(∼)} × K & ↪ {\closSim} \\  
    0 & ↦ (0, 0) \\
    1 & ↦ (1, 1) \\
    (s, t), p & ↦ (s, p), (t, p) 
\end{align*}
These two embeddings allows us to construct the following pullback square:
\[
    \begin{tikzcd}
        \norm(∼) \ar[hook]{r}{i} \ar[dashed, swap]{d}{\norm(h)} & ∼ \ar{d}{h}\\  
        G(\norm(∼)) \ar[hook]{r}{i} & (\closSim)^\At
    \end{tikzcd}
\]
